%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%\part{Index}
%   \printindex
%\part{Appendix}
\mychapter{Mathematica notebooks.}
   \input{mathematica.tex}
\mychapter{Spherical N-pendulum problem.}
   \input{mine/multiPendulumSphericalMatrix.tex}
%
%\part{Lagrangian Topics.}
   \mychapter{Newton's Law from Lagrangian.}
      \input{galagrangian/newtonianLagrangianAndGradient.tex}  % Aug 9/08
   \mychapter{Covariant Lagrangian, and electrodynamic potential.}
      \input{galagrangian/srLagrangian.tex}                      % Aug 21/08
   \mychapter{Vector canonical momentum.}
      \input{galagrangian/canonicalMomentum.tex}                 % Sep 1/08
   \mychapter{Direct variation of Maxwell equations.}
      \input{galagrangian/lagrangianFieldDensity.tex}           % Sep 8/08
   \mychapter{Revisit Lorentz force from Lagrangian.}
      \input{galagrangian/lorentzForce.tex}                      % Oct 8/08
   \mychapter{Tensor Derivation of Covariant Lorentz Force from Lagrangian.}
      \input{galagrangian/maxwellTensorLagrangian.tex}          % Oct 12/08
   \mychapter{Lorentz Invariance of Maxwell Lagrangian.}
      \input{galagrangian/boostMaxwellLagrangian.tex}           % Oct 19/08
   \mychapter{Lorentz transform Noether current (interaction Lagrangian).}
      \input{galagrangian/noethersLorentzForce.tex}             % Oct 22/08
   \mychapter{Lorentz force Lagrangian with conjugate momentum.}
      \input{galagrangian/lorentzForcePQA.tex}                 % Apr 15/09
   \mychapter{Tensor Maxwell equation.}
      \input{galagrangian/maxwellTensorFromLagrangian.tex}     % Apr 20/09
   \mychapter{Canonical energy momentum tensor and translation.}
      \input{galagrangian/stressEnergyNoethers.tex}
   \mychapter{Covariant Lorentz force Lagrangians.}
      \input{galagrangian/lForceLag2.tex}
   \mychapter{Lorentz force from Lagrangian (non-covariant.)}
      \input{galagrangian/jackson12Dash9.tex}
   \mychapter{Spherical polar pendulum chain (multivector.)}
      \input{galagrangian/sPolarMultiPendulum.tex}
   \mychapter{Spherical polar pendulum (take II.)}
      \input{galagrangian/multiPendulumSpherical2.tex}
%
%\part{Chronology}
%   \include{cronology}
