%
% Copyright © 2013 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
\part{Organizational framework.}
   \mychapter{Potential and Kinetic energy.}
      \input{mine/potentialEnergy.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Calculus of variations.}
      \input{mine/goldsteinCh12ShortestCurveVariational.tex} % p2.1
      \input{mine/goldsteinCh12GeodesicsOnSphere.tex} % p2.2
      \input{mine/goldsteinCh12EulerLagrange2ndOrder.tex} % p2.4
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Special relativity.}
      \paragraph{Placeholder.}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Action and Euler-Langrange equations.}
      \input{galagrangian/eulerLagrangeScalarLagrangian.tex}
      \section{Problems.}
         \input{problems/classicalMechanicsPs1LorentzForce.tex} % p1.1
         \input{problems/classicalMechanicsPs1BallThrown.tex} % p1.2
         \input{problems/classicalMechanicsPs1ChangeOfVariables.tex} % p1.3
         \input{problems/classicalMechanicsPs2MotionInEMfields.tex} % p2.1
         \input{problems/classicalMechanicsPs2AngularMomentumInSphericalAndCylCoords.tex} % p2.5
         \input{mine/goldsteinCh12AngularMomentumThreeParticle.tex} % p1.8
         \input{mine/goldsteinCh12KEforBarbell.tex} % p1.6
         \input{galagrangian/tongMf1PurelyKinetic.tex} % p1
         \input{galagrangian/tongMf1AlternateLagrangian.tex} % p2
         \input{galagrangian/tongMf1RelativisticEOM.tex} % p3
         \input{galagrangian/tongMf1DoublePendulum.tex} % p4
         \input{galagrangian/tongMf1LorentzForceLagrangian.tex} % p6
         \input{problems/dipoleMomentConstantElectricField.tex}
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Constraints.}
      \input{galagrangian/tongMf1PendulumOnRotatingWheel.tex} % p5
      \input{galagrangian/tongMf1CircularlyConstrainedPaths.tex} % p7
      \input{galagrangian/tongMf1MassesOnStringOneDangling.tex} % p8
      \input{problems/landauMechanicsCh1P3SupportMovingInCircle.tex} % p1.3
      \input{problems/landauMechanicsCh1P3PendulumSupportInMotion.tex} % p1.3b
      \input{problems/landauMechanicsCh1P3PendulumVerticalSupportMotion.tex} % p1.3c
      \input{mine/hoopSpring.tex}
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Geometric algebra.}
      \paragraph{Placeholder.}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Noether's theorem.}
      \input{galagrangian/eulerLagrange.tex}                     % Oct 13/08
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Hamiltonian mechanics.}
      \input{mine/hamiltonian.tex}
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Routhian procedure.}
      \input{mine/goldsteinRouth.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Rigid body motion.}
      \input{lectures/rigidBodyMotion.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Euler angles.}
      \input{lectures/classicalMechanicsEulerAngles.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Parallel axis theorem.}
      \input{lectures/parallelAxisTheorem.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Phase Space and Trajectories.}
      \input{lectures/phaseSpaceAndTrajectories.tex}
      %\section{Solutions.}
      %   \shipoutAnswer
   \mychapter{Conserved quantities.}
      \section{Runge-Lenz vector conservation.}
         \input{lectures/RungeLenz.tex}
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Field Lagrangians.}
      \input{galagrangian/fieldLagrangian.tex}                   % Oct 10/08
      \input{mine/waveLagrangianOneDimension.tex}
      \input{mine/waveLagrangianMoreDimensions.tex}
      \input{mine/waveLagrangianQM.tex}
      \input{mine/waveLagrangianKG.tex}
      \input{mine/waveLagrangianDirac.tex}
      \section{Solutions.}
         \shipoutAnswer
   \mychapter{Noether's theorem for fields.}
      \section{Noether's theorem.}
         \input{galagrangian/noethersField.tex}                     % Oct 29/08
      \section{Translation and rotation Noether field currents.}
         \input{galagrangian/rotationCurrents.tex}
      \section{Solutions.}
         \shipoutAnswer
%
\part{Lagrangian Topics.}
   \mychapter{Newton's Law from Lagrangian.}
      \input{galagrangian/newtonianLagrangianAndGradient.tex}  % Aug 9/08
   \mychapter{Covariant Lagrangian, and electrodynamic potential.}
      \input{galagrangian/srLagrangian.tex}                      % Aug 21/08
   \mychapter{Vector canonical momentum.}
      \input{galagrangian/canonicalMomentum.tex}                 % Sep 1/08
   \mychapter{Direct variation of Maxwell equations.}
      \input{galagrangian/lagrangianFieldDensity.tex}           % Sep 8/08
   \mychapter{Revisit Lorentz force from Lagrangian.}
      \input{galagrangian/lorentzForce.tex}                      % Oct 8/08
   \mychapter{Tensor Derivation of Covariant Lorentz Force from Lagrangian.}
      \input{galagrangian/maxwellTensorLagrangian.tex}          % Oct 12/08
   \mychapter{Lorentz Invariance of Maxwell Lagrangian.}
      \input{galagrangian/boostMaxwellLagrangian.tex}           % Oct 19/08
   \mychapter{Lorentz transform Noether current (interaction Lagrangian).}
      \input{galagrangian/noethersLorentzForce.tex}             % Oct 22/08
   \mychapter{Lorentz force Lagrangian with conjugate momentum.}
      \input{galagrangian/lorentzForcePQA.tex}                 % Apr 15/09
   \mychapter{Tensor Maxwell equation.}
      \input{galagrangian/maxwellTensorFromLagrangian.tex}     % Apr 20/09
   \mychapter{Canonical energy momentum tensor and translation.}
      \input{galagrangian/stressEnergyNoethers.tex}
   \mychapter{Covariant Lorentz force Lagrangians.}
      \input{galagrangian/lForceLag2.tex}
   \mychapter{Lorentz force from Lagrangian (non-covariant.)}
      \input{galagrangian/jackson12Dash9.tex}
   \mychapter{Spherical polar pendulum chain (multivector.)}
      \input{galagrangian/sPolarMultiPendulum.tex}
   \mychapter{Spherical polar pendulum (take II.)}
      \input{galagrangian/multiPendulumSpherical2.tex}
%
\part{Random independent study notes.}
   \mychapter{Infinite homogeneous plane potential.}
      \input{mine/infiniteSheetPotentials.tex}
   \mychapter{Infinitesimal width infinite plane.}
      \input{mine/InfinitePlanePotentialTakeIII.tex}
   \mychapter{Potential due to cylindrical distribution.}
      \input{mine/cylinderPotential.tex}
