%
% Copyright © 2012 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{blah}
\renewcommand{\dirname}{notes/classicalmechanics/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../latex/peeter_prologue_print2.tex}

%\usepackage{peeters_layout_exercise}
%\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{macros_cal}

\newcommand{\nbref}[1]{%
\itemRef{classicalmechanics}{#1}%
\index{Mathematica}%
}

\newcommand{\bookchapcite}[2]{\ref{chap:#1}}
\newcommand{\chapcite}[1]{\ref{chap:#1}}

\beginArtNoToc

\generatetitle{Refuse.}

\chapter{Compare some wave equation's and their Lagrangians.}
   \input{.junk/waveLagrangian.tex}
\chapter{Short metric tensor explanation.}
   \input{.junk/lorentzMetricTensor.tex}
\chapter{Two by two positive definite invariant tx.}
    \input{.junk/quadraticForm.tex}
\chapter{1D forced harmonic oscillator (non-homogeneous.)}
    \input{.junk/1dharmonicOsc.tex}
\chapter{EOM for a one dimensional problem.}
    \input{.junk/1dpotentialIntegral.tex}
\chapter{Infinite homogeneous plane potential.}
   \input{.junk/infiniteSheetPotentials.tex}
\chapter{Infinitesimal width infinite plane.}
   \input{.junk/InfinitePlanePotentialTakeIII.tex}
\chapter{Potential due to cylindrical distribution.}
   \input{.junk/cylinderPotential.tex}
\chapter{Newton's Law from Lagrangian.}
   \input{.junk/newtonianLagrangianAndGradient.tex}  % Aug 9/08
\chapter{Covariant Lorentz force Lagrangians.}
   \input{.junk/lForceLag2.tex}
\chapter{Covariant Lagrangian, and electrodynamic potential.}
   \input{.junk/srLagrangian.tex}                      % Aug 21/08
\chapter{Revisit Lorentz force from Lagrangian.}
   \input{.junk/lorentzForce.tex}                      % Oct 8/08
\chapter{Tensor Derivation of Covariant Lorentz Force from Lagrangian.}
   \input{.junk/maxwellTensorLagrangian.tex}          % Oct 12/08
\chapter{Lorentz force Lagrangian with conjugate momentum.}
   \input{.junk/lorentzForcePQA.tex}                 % Apr 15/09
\chapter{Spherical polar pendulum chain (multivector.)}
   \input{.junk/sPolarMultiPendulum.tex}
\chapter{Spherical polar pendulum (take II.)}
   \input{.junk/multiPendulumSpherical2.tex}
\chapter{Random purged problems.}
   \input{.junk/tongMf1LorentzForceLagrangianJunk.tex}
\chapter{Maybe interesting: alternate covariant kinetic Lagrangian.}
   \input{maybeInterestingAlternateCovariantKineticLagrangian.tex}

%}
%\EndArticle
\EndNoBibArticle
