%
% Copyright © 2020 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{maxwells}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{notes/ece1228-electromagnetic-theory/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../latex/peeter_prologue_print2.tex}

\PassOptionsToPackage{answerdelayed}{exercise}

% proof:
\usepackage{amsthm}
\usepackage{macros_cal} % \LL
\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
\usepackage{verbatim}
%\usepackage{mhchem} % \ce{}
%\usepackage{macros_bm} % \bcM
%\usepackage{macros_qed} % \qedmarker
%\usepackage{txfonts} % \ointclockwise

\beginArtNoToc

\generatetitle{Maxwell's equation using geometric algebra Lagrangian.}
%\chapter{XXX}
%\label{chap:maxwells}
\section{Motivation.}
In my classical mechanics notes, I've got computations of Maxwell's equation (singular in it's geometric algebra form) from a Lagrangian in various ways (using a tensor, scalar and multivector Lagrangians), but all of these seem more convoluted than they should be.  Here's a new derivation, starting with the action principle for field variables.
\section{Field action.}
\maketheorem{Relativistic Euler-Lagrange field equations.}{thm:maxwells:40}{
Let \( \phi \rightarrow \phi + \delta \phi \) be any variation of the field, such that the variation
\( \delta \phi = 0 \) vanishes at the boundaries of the action integral
\begin{equation*}
S = \int d^4 x \LL(\phi, \partial_\nu \phi).
\end{equation*}
The extreme value of the action is found when the Euler-Lagrange equations
\begin{equation*}
0 = \PD{\phi}{\LL} - \partial_\nu \PD{(\partial_\nu \phi)}{\LL},
\end{equation*}
are satisfied.  For a Lagrangian with multiple field variables, there will be one such equation for each field.
} % theorem
\begin{proof}
To ease the visual burden, designate the variation of the field by \( \delta \phi = \epsilon \), and perform a first order expansion of the varied Lagrangian
\begin{dmath}\label{eqn:maxwells:20}
\LL \rightarrow
\LL(\phi + \epsilon, \partial_\nu (\phi + \epsilon))
=
\LL(\phi, \partial_\nu \phi)
+
\PD{\phi}{\LL} \epsilon +
\PD{(\partial_\nu \phi)}{\LL} \partial_\nu \epsilon.
\end{dmath}
The variation of the Lagrangian is
\begin{dmath}\label{eqn:maxwells:40}
\delta \LL =
\PD{\phi}{\LL} \epsilon +
\PD{(\partial_\nu \phi)}{\LL} \partial_\nu \epsilon
=
\PD{\phi}{\LL} \epsilon +
\partial_\nu \lr{ \PD{(\partial_\nu \phi)}{\LL} \epsilon }
-
\epsilon \partial_\nu \PD{(\partial_\nu \phi)}{\LL},
\end{dmath}
which we may plug into the action integral to find
\begin{dmath}\label{eqn:maxwells:60}
   \delta S =
   \int d^4 x \epsilon \lr{
   \PD{\phi}{\LL}
   -
   \partial_\nu \PD{(\partial_\nu \phi)}{\LL}
}
+
   \int d^4 x
\partial_\nu \lr{ \PD{(\partial_\nu \phi)}{\LL} \epsilon }.
\end{dmath}
The last integral can be evaluated along the \( dx^\nu \) direction, leaving
\begin{dmath}\label{eqn:maxwells:80}
   \int d^3 x
   \evalbar{ \PD{(\partial_\nu \phi)}{\LL} \epsilon }{\Delta x^\nu},
\end{dmath}
where \( d^3 x = dx^\alpha dx^\beta dx^\gamma \) is the product of differentials that does not include \( dx^\nu \).  By construction, \( \epsilon \) vanishes on the boundary of the action integral so \cref{eqn:maxwells:80} is zero.  The action takes its extreme value when
\begin{dmath}\label{eqn:maxwells:100}
0 = \delta S
=
\int d^4 x \epsilon \lr{
   \PD{\phi}{\LL}
   -
   \partial_\nu \PD{(\partial_\nu \phi)}{\LL}
}.
\end{dmath}
The proof is complete after noting that this must hold for all variations of the field \( \epsilon \), which means that we must have
\begin{dmath}\label{eqn:maxwells:120}
0  =
   \PD{\phi}{\LL}
   -
   \partial_\nu \PD{(\partial_\nu \phi)}{\LL}.
\end{dmath}
\end{proof}
%}
Armed with the Euler-Lagrange equations, we can apply them to the Maxwell's equation Lagrangian, which we will claim has the following form.
\maketheorem{Maxwell's equation Lagrangian.}{thm:maxwells:140}{
Application of the Euler-Lagrange equations to the Lagrangian
\begin{equation*}
\LL = - \frac{\epsilon_0 c}{2} F \cdot F + J \cdot A,
\end{equation*}
where \( F = \grad \wedge A \), yields the vector portion of Maxwell's equation
\begin{equation*}
\grad \cdot F = \inv{\epsilon_0 c} J,
\end{equation*}
which implies
\begin{equation*}
\grad F = \inv{\epsilon_0 c} J.
\end{equation*}
This is Maxwell's equation.
} % definition
\begin{proof}
We wish to apply all of the Euler-Lagrange equations simulaneously (i.e. once for each of the four \(A_\mu\) components of the potential), and cast it into four-vector form
\begin{dmath}\label{eqn:maxwells:140}
0 = \gamma_\nu \lr{ \PD{A_\nu}{} - \partial_\mu \PD{(\partial_\mu A_\nu)}{} } \LL.
\end{dmath}
Since our Lagrangian splits nicely into kinetic and interaction terms, this gives us
\begin{dmath}\label{eqn:maxwells:160}
   0 = \gamma_\nu \lr{ \PD{A_\nu}{(A \cdot J)} + \frac{\epsilon_0 c}{2} \partial_\mu \PD{(\partial_\mu A_\nu)}{ (F \cdot F)} }.
\end{dmath}
The interaction term above is just
\begin{equation}\label{eqn:maxwells:180}
   \gamma_\nu \PD{A_\nu}{(A \cdot J)}
=
\gamma_\nu \PD{A_\nu}{(A_\mu J^\mu)}
=
\gamma_\nu J^\nu
=
J,
\end{equation}
but the kinetic term takes a bit more work.  Let's start with evaluating
\begin{dmath}\label{eqn:maxwells:200}
   \PD{(\partial_\mu A_\nu)}{ (F \cdot F)}
=
\PD{(\partial_\mu A_\nu)}{ F } \cdot F
+
F \cdot \PD{(\partial_\mu A_\nu)}{ F }
=
2 \PD{(\partial_\mu A_\nu)}{ F } \cdot F
=
2 \PD{(\partial_\mu A_\nu)}{ (\partial_\alpha A_\beta) } \lr{ \gamma^\alpha \wedge \gamma^\beta } \cdot F
=
2 \lr{ \gamma^\mu \wedge \gamma^\nu } \cdot F.
\end{dmath}
We hit this with the mu-partial and expand as a scalar selection to find
\begin{dmath}\label{eqn:maxwells:220}
   \partial_\mu \PD{(\partial_\mu A_\nu)}{ (F \cdot F)}
=
2 \lr{ \partial_\mu \gamma^\mu \wedge \gamma^\nu } \cdot F
=
- 2 (\gamma^\nu \wedge \grad) \cdot F
=
- 2 \gpgradezero{ (\gamma^\nu \wedge \grad) F }
=
- 2 \gpgradezero{ \gamma^\nu \grad F - \cancel{ \gamma^\nu \cdot \grad F } }
=
- 2 \gamma^\nu \cdot \lr{ \grad \cdot F }.
\end{dmath}
Putting all the pieces together yields
\begin{equation}\label{eqn:maxwells:240}
0
= J - \epsilon_0 c \gamma_\nu \lr{ \gamma^\nu \cdot \lr{ \grad \cdot F } }
= J - \epsilon_0 c \lr{ \grad \cdot F },
\end{equation}
but
\begin{dmath}\label{eqn:maxwells:260}
\grad \cdot F
=
\grad F - \grad \wedge F
=
\grad F - \grad \wedge (\grad \wedge A)
=
\grad F,
\end{dmath}
so the multivector field equations for this Lagrangian are
\begin{dmath}\label{eqn:maxwells:280}
\grad F = \inv{\epsilon_0 c} J,
\end{dmath}
as claimed.
\end{proof}
\makeproblem{Correspondence with tensor formalism.}{problem:maxwells:300}{
Cast the Lagrangian of \cref{thm:maxwells:140} into the conventional tensor form
\begin{dmath}\label{eqn:maxwells:300}
\LL = \frac{\epsilon_0 c}{4} F_{\mu\nu} F^{\mu\nu} + A^\mu J_\mu.
\end{dmath}
Also show that the four-vector component of Maxwell's equation \( \grad \cdot F = J/(\epsilon_0 c \) is equivalent to the conventional tensor form of the Gauss-Ampere law
\begin{dmath}\label{eqn:maxwells:320}
\partial_\mu F^{\mu\nu} = \inv{\epsilon_0 c} J^\nu,
\end{dmath}
where \( F^{\mu\nu} = \partial^\mu A^\nu - \partial^\nu A^\mu \) as usual.  Also
show that the trivector component of Maxwell's equation
\( \grad \wedge F =  0 \) is equivalent to the tensor form of the Gauss-Faraday law
\begin{dmath}\label{eqn:maxwells:340}
\partial_\alpha \lr{ \epsilon^{\alpha \beta \mu \nu} F_{\mu\nu} } = 0.
\end{dmath}
} % problem
\makeanswer{problem:maxwells:300}{
To show the Lagrangian correspondence we must expand \( F \cdot F \) in coordinates
\begin{dmath}\label{eqn:maxwells:360}
F \cdot F
=
( \grad \wedge A ) \cdot
( \grad \wedge A )
=
\lr{ (\gamma^\mu \partial_\mu) \wedge (\gamma^\nu A_\nu) }
\cdot
\lr{ (\gamma^\alpha \partial_\alpha) \wedge (\gamma^\beta A_\beta) }
=
\lr{ \gamma^\mu \wedge \gamma^\nu } \cdot \lr{ \gamma_\alpha \wedge \gamma_\beta }
(\partial_\mu A_\nu )
(\partial^\alpha A^\beta )
=
\lr{
{\delta^\mu}_\beta
{\delta^\nu}_\alpha
-
{\delta^\mu}_\alpha
{\delta^\nu}_\beta
}
(\partial_\mu A_\nu )
(\partial^\alpha A^\beta )
=
- \partial_\mu A_\nu \lr{
   \partial^\mu A^\nu
   -
   \partial^\nu A^\mu
}
=
- \partial_\mu A_\nu F^{\mu\nu}
=
- \inv{2} \lr{
   \partial_\mu A_\nu F^{\mu\nu}
   +
   \partial_\nu A_\mu F^{\nu\mu}
}
=
- \inv{2} \lr{
   \partial_\mu A_\nu
   -
   \partial_\nu A_\mu
}
F^{\mu\nu}
=
-
\inv{2}
F_{\mu\nu}
F^{\mu\nu}.
\end{dmath}
With a substitution of this and \( A \cdot J = A_\mu J^\mu \) back into the Lagrangian, we recover the tensor form of the Lagrangian.

To recover the tensor form of Maxwell's equation, we first split it into vector and trivector parts
\begin{dmath}\label{eqn:maxwells:1580}
\grad \cdot F + \grad \wedge F = \inv{\epsilon_0 c} J.
\end{dmath}
Now the vector component may be expanded in coordinates by dotting both sides with \( \gamma^\nu \) to find
\begin{dmath}\label{eqn:maxwells:1600}
\inv{\epsilon_0 c} \gamma^\nu \cdot J = J^\nu,
\end{dmath}
and
\begin{dmath}\label{eqn:maxwells:1620}
\gamma^\nu \cdot
\lr{ \grad \cdot F }
=
\partial_\mu \gamma^\nu \cdot \lr{ \gamma^\mu \cdot \lr{ \gamma_\alpha \wedge \gamma_\beta } \partial^\alpha A^\beta }
=
\lr{
   {\delta^\mu}_\alpha
   {\delta^\nu}_\beta
   -
   {\delta^\nu}_\alpha
   {\delta^\mu}_\beta
}
\partial_\mu
\partial^\alpha A^\beta
=
\partial_\mu
\lr{
   \partial^\mu A^\nu
   -
   \partial^\nu A^\mu
}
=
\partial_\mu F^{\mu\nu}.
\end{dmath}
Equating \cref{eqn:maxwells:1600} and \cref{eqn:maxwells:1620} finishes the first part of the job.  For the trivector component, we have
\begin{dmath}\label{eqn:maxwells:1640}
0
= \grad \wedge F
= (\gamma^\mu \partial_\mu) \wedge \lr{ \gamma^\alpha \wedge \gamma^\beta } \partial_\alpha A_\beta
= \inv{2} (\gamma^\mu \partial_\mu) \wedge \lr{ \gamma^\alpha \wedge \gamma^\beta } F_{\alpha \beta}.
\end{dmath}
Wedging with \( \gamma^\tau \) and then multiplying by \( -2 I \) we find
\begin{dmath}\label{eqn:maxwells:1660}
0 = - \lr{ \gamma^\mu \wedge \gamma^\alpha \wedge \gamma^\beta \wedge \gamma^\tau } I \partial_\mu F_{\alpha \beta},
\end{dmath}
but
\begin{dmath}\label{eqn:maxwells:1680}
\gamma^\mu \wedge \gamma^\alpha \wedge \gamma^\beta \wedge \gamma^\tau = -I \epsilon^{\mu \alpha \beta \tau},
\end{dmath}
which leaves us with
\begin{dmath}\label{eqn:maxwells:1700}
\epsilon^{\mu \alpha \beta \tau}  \partial_\mu F_{\alpha \beta} = 0,
\end{dmath}
as expected.
} % answer
\makeproblem{Correspondence of tensor and Gibbs forms of Maxwell's equations.}{problem:maxwells:379}{
Given the identifications
%\cref{eqn:lorentzForceCovariant:1540}
\begin{subequations}
\label{eqn:lorentzForceCovariant:1540}
\begin{equation}\label{eqn:lorentzForceCovariant:1500}
F^{k0} = E^k,
\end{equation}
and
\begin{equation}\label{eqn:lorentzForceCovariant:1520}
F^{rs} = -\epsilon^{rst} B^t,
\end{equation}
\end{subequations}
and
\begin{dmath}\label{eqn:maxwells:1560}
J^\mu = \lr{ c \rho, \BJ },
\end{dmath}
the reader should satisfy themselves that the traditional Gibbs form of Maxwell's equations can be recovered from \cref{eqn:maxwells:320}.
} % problem
\makeanswer{problem:maxwells:379}{
The reader is referred to Exercise 3.4 ``Electrodynamics, variational principle.'' from \citep{pjootQFT}.
} % answer
\makeproblem{Correspondence with grad and curl form of Maxwell's equations.}{problem:maxwells:380}{
With \( J = c \rho \gamma_0 + J^k \gamma_k \) and \( F = \BE + I c \BB \) show that Maxwell's equation, as stated in \cref{thm:maxwells:140} expand to the conventional div and curl expressions for Maxwell's equations.
} % problem
\makeanswer{problem:maxwells:380}{
To obtain Maxwell's equations in their traditional vector forms, we pre-multiply both sides with \( \gamma_0 \)
\begin{dmath}\label{eqn:maxwells:1720}
\gamma_0 \grad F = \inv{\epsilon_0 c} \gamma_0 J,
\end{dmath}
and then select each grade separately.  First observe that the RHS above has scalar and bivector components, as
\begin{dmath}\label{eqn:maxwells:1740}
\gamma_0 J
=
c \rho + J^k \gamma_0 \gamma_k.
\end{dmath}
In terms of the spatial bivector basis \( \Be_k = \gamma_k \gamma_0 \), the RHS of \cref{eqn:maxwells:1720} is
\begin{dmath}\label{eqn:maxwells:1760}
\gamma_0 \frac{J}{\epsilon_0 c} = \frac{\rho}{\epsilon_0} - \mu_0 c \BJ.
\end{dmath}
For the LHS, first note that
\begin{dmath}\label{eqn:maxwells:1780}
\gamma_0 \grad
=
\gamma_0
\lr{
   \gamma_0 \partial^0 +
   \gamma_k \partial^k
}
=
\partial_0 - \gamma_0 \gamma_k \partial_k
=
\inv{c} \PD{t}{} + \spacegrad.
\end{dmath}
We can express all the the LHS of \cref{eqn:maxwells:1720} in the bivector spatial basis, so that Maxwell's equation in multivector form is
\begin{dmath}\label{eqn:maxwells:1800}
\lr{ \inv{c} \PD{t}{} + \spacegrad } \lr{ \BE + I c \BB } = \frac{\rho}{\epsilon_0} - \mu_0 c \BJ.
\end{dmath}
Selecting the scalar, vector, bivector, and trivector grades of both sides (in the spatial basis) gives the following set of respective equations
\begin{subequations}
\label{eqn:maxwells:1820}
\begin{dmath}\label{eqn:maxwells:1840}
\spacegrad \cdot \BE = \frac{\rho}{\epsilon_0}
\end{dmath}
\begin{dmath}\label{eqn:maxwells:1860}
\inv{c} \partial_t \BE + I c \spacegrad \wedge \BB = - \mu_0 c \BJ
\end{dmath}
\begin{dmath}\label{eqn:maxwells:1880}
\spacegrad \wedge \BE + I \partial_t \BB = 0
\end{dmath}
\begin{dmath}\label{eqn:maxwells:1900}
I c \spacegrad \cdot B = 0,
\end{dmath}
\end{subequations}
which we can rewrite after some duality transformations (and noting that \( \mu_0 \epsilon_0 c^2 = 1 \)), we have
\begin{subequations}
\label{eqn:maxwells:1920}
\begin{dmath}\label{eqn:maxwells:1940}
\spacegrad \cdot \BE = \frac{\rho}{\epsilon_0}
\end{dmath}
\begin{dmath}\label{eqn:maxwells:1960}
\spacegrad \cross \BB - \mu_0 \epsilon_0 \PD{t}{\BE} = \mu_0 \BJ
\end{dmath}
\begin{dmath}\label{eqn:maxwells:1980}
\spacegrad \cross \BE + \PD{t}{\BB} = 0
\end{dmath}
\begin{dmath}\label{eqn:maxwells:2000}
\spacegrad \cdot B = 0,
\end{dmath}
\end{subequations}
which are Maxwell's equations in their traditional form.
} % answer
\subsection{Solutions.}
\shipoutAnswer
\EndArticle
