%
% Copyright © 2020 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{lorentzForceCovariant}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{notes/ece1228-electromagnetic-theory/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../latex/peeter_prologue_print2.tex}

\PassOptionsToPackage{answerdelayed}{exercise}

% proof:
\usepackage{amsthm}

%\usepackage[thmmarks]{ntheorem}
%\theoremheaderfont{\bfseries}
%\theorembodyfont{\normalfont}
%\theoremseparator{}
%\theoremsymbol{$\blacksquare$}
%\newtheorem*{answerproof}{}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
\usepackage{verbatim}
%\usepackage{mhchem} % \ce{}
%\usepackage{macros_bm} % \bcM
%\usepackage{macros_qed} % \qedmarker
%\usepackage{txfonts} % \ointclockwise

\beginArtNoToc

\generatetitle{Lagrangian for the Lorentz force equation.}
\section{Motivation.}
In my old classical mechanics notes it appears that I did covariant derivations of the Lorentz force equations a number of times, using different trial Lagrangians (relativistic and non-relativistic), and using both geometric algebra and tensor methods.
However, none of these appear to have been done concisely, and a number not even coherently.

The following document has been drafted as replacement text for those incoherent classical mechanics notes.
I'll attempt to cover
\begin{itemize}
\item a lighting review of the geometric algebra STA (Space Time Algebra),
\item relations between Dirac matrix algebra and STA,
\item derivation of the relativistic form of the Euler-Lagrange equations from the covariant form of the action,
\item relationship of the STA form of the Euler-Lagrange equations to their tensor equivalents,
\item derivation of the Lorentz force equation from the STA Lorentz force Lagrangian,
\item relationship of the STA Lorentz force equation to its equivalent in the tensor formalism,
\item relationship of the STA Lorentz force equation to the traditional vector form.
\end{itemize}
Note that some of the prerequisite ideas and auxiliary details are presented as problems with solutions.
If the reader has sufficient background to attempt those problems themselves, they are encouraged to do so.

\input{sta.tex}
\input{covariantSTA.tex}

\subsection{Solutions.}
\shipoutAnswer
%}
\EndArticle
%\EndNoBibArticle
